---
title: "wvs"
author: "Marek Chadim"
date: "2024-09-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(countrycode)

rm(list=ls())
gc()
load("C:/Users/chadi/Dropbox/Applied-Economics/WVS/WV6_Data_R_v20201117.rdata")
```

# 1 Download data and documentation

The dataset and the codebook from the World Value Survey can be
downloaded from
<http://www.worldvaluessurvey.org/WVSDocumentationWV6.jsp>

The data has $89{,}565$ rows and $442$ columns.

```{r}
dim(WV6_Data_R_v20201117)
```

# 2 Creating the data set

```{r}
# Create the country variable based on country codes
d <- WV6_Data_R_v20201117
d <- d |> mutate(country = countrycode(V2, origin = "iso3n", destination = "country.name"))

# View the country variable to verify correct labeling
View(d$country)

# Select variables related to values, social trust, happiness, and political orientation
selected_vars <- c("V5", "V6", "V7", "V8", "V9", "V10", "V11", "V23", "V24", 
                   "V55", "V56", "V59", "V67", "V69", "V70", "V71", "V72", 
                   "V73", "V74", "V76", "V77", "V78", "V79", "V102", "V103", 
                   "V104", "V105", "V106", "V107", "V108", "V109", "V110", 
                   "V111", "V112", "V113", "V114", "V115", "V116", "V117", 
                   "V118", "V119", "V120", "V121", "V122", "V123", "V124", 
                   "V127", "V128", "V130", "V131", "V132", "V133", "V134", 
                   "V135", "V136", "V137", "V138", "V139", "V97", "V98", 
                   "V99", "V100", "V101", "V140", "V141", "V142", "V143", 
                   "V145", "V146", "V147", "V148", "V149", "V150", "V151", 
                   "V152", "V153", "V154", "V155", "V156", "V157", "V158", 
                   "V159", "V160", "V161", "V162", "V163", "V164", "V165", 
                   "V166", "V167", "V168", "V169", "V170")

# Subset the data to the selected variables
d <- d |> select(country, all_of(selected_vars))

# Handle missing values by replacing specific negative codes with NA
d <- d |> 
  mutate(across(where(is.numeric), ~na_if(., -1))) |>
  mutate(across(where(is.numeric), ~na_if(., -2))) |>
  mutate(across(where(is.numeric), ~na_if(., -3))) |>
  mutate(across(where(is.numeric), ~na_if(., -4))) |>
  mutate(across(where(is.numeric), ~na_if(., -5)))

# Aggregate data to the country level by calculating the mean of all variables
d <- d |> group_by(country) |> summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)))

# Remove any columns with missing data
d <- d[, colSums(is.na(d)) == 0]
colSums(is.na(d)) == 0  # Check to confirm no missing values remain

# Write the cleaned data to a CSV file
write.csv(d, "wvs.csv", row.names = FALSE)
```

# 3 Analysis

## PCA

PCA is performed on the dataset after scaling the variables to have a
standard deviation of one.

```{r}
wvs <- read.csv("wvs.csv")
row.names(wvs) <- wvs[,1]
wvs <- wvs[,-1]   
countries <- row.names(wvs)
countries
names(wvs)
```

Without normalization, results differ due to variable scale differences:

```{r}
apply(wvs, 2, mean)
apply(wvs, 2, var)
```

Running PCA with scaling:

```{r}
pr.out <- prcomp(wvs, scale = TRUE)
names(pr.out)
dim(pr.out$x)
summary(pr.out)
plot(pr.out)
```

A scree plot shows the PVE by each principal component, with the first
four explaining over 50% of the variance:

```{r chunk49}
pve <- 100 * pr.out$sdev^2 / sum(pr.out$sdev^2)
par(mfrow = c(1, 2))
plot(pve,  type = "o", ylab = "PVE",
    xlab = "Principal Component", col = "blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE",
    xlab = "Principal Component", col = "brown3")
```

## Cultural Map

A biplot visualizes countries in the principal component space:

```{r}
par(mfrow = c(1, 1))
biplot(pr.out, scale = 0, cex=.5)
```

### Interpretation

PC1 (Self-expression vs. Survival values): This axis likely reflects the
degree to which countries emphasize individual freedoms,
self-expression, and creativity (variables such as V70 "New ideas and
creativity"). Countries scoring high on this axis, like Sweden and
Germany, represent liberal democracies with a high emphasis on
self-expression and secular values. In contrast, countries scoring
lower, such as Uzbekistan and Qatar, tend to prioritize survival values
such as tradition and authority .

PC2 (Economic vs. Traditional values): The second component seems to be
linked to economic ideology and traditional beliefs. Variables such as
V132 ("Religious authorities interpret laws") and V97 ("Private vs.
state ownership") load heavily on this component, indicating a spectrum
from market-driven economies to more traditionally governed societies.
Japan and Sweden score high on this component, reflecting their advanced
economies and progressive values, while countries like Haiti and Egypt
cluster together, possibly due to their emphasis on traditional
authority structures

The analysis confirms the importance of self-expression versus survival
values along the first principal component (PC1). However, the second
principal component (PC2) does not clearly align with the
secular-rational versus traditional divide. Instead, it seems to capture
a spectrum of social hierarchy, religious authority, and respect for
governance. Countries like Haiti and Egypt score high on PC2, reflecting
the role of religious influence and social obedience, whereas countries
like Japan and Sweden score lower, signaling a preference for secular,
rational governance with less hierarchical structures.

## Clustering Analysis of Countries

```{r}
wvs <- wvs[,-1]     
sd.data <- scale(wvs)
hc.out <- hclust(dist(sd.data))
hc.clusters <- cutree(hc.out, 4)
table(hc.clusters, row.names(wvs))
```

```{r}
par(mfrow = c(1, 1))
plot(hc.out, labels = row.names(wvs))
abline(h = 139, col = "red")
```

### Interpretation of Country Clusters:

Distinct Cluster - Haiti: Haiti stands out at a high level in the hierarchy, suggesting its responses are notably distinct, likely due to unique social, political, or cultural traits.

Western Countries: The USA, Germany, Sweden, New Zealand, Australia, and the Netherlands form a close cluster, reflecting high-income, industrialized nations with shared values of individualism, democracy, secularism, and human rights.

Latin American and European Countries: Argentina, Brazil, Chile, Peru, Spain, and Slovenia cluster together, likely blending European heritage with Latin American values focused on community, family, and moderate economic priorities.

Middle Eastern and African Countries: Pakistan, Kuwait, Iraq, Libya, South Africa, Nigeria, and Malaysia form a broader cluster, reflecting traditionalism, strong family values, and religious orientations.

Central/Eastern Europe and Asia: Russia, Kazakhstan, Azerbaijan, and China cluster, possibly due to post-Soviet or authoritarian influences, sharing similar values around authority and governance.

## Clustering Analysis of Variables

```{r}
wvs.data <- t(wvs)
sd.data <- scale(wvs.data)
hc.out <- hclust(dist(sd.data))
hc.clusters <- cutree(hc.out, 4)
table(hc.clusters, colnames(wvs))
```

```{r}
par(mfrow = c(1, 1))
plot(hc.out, labels = colnames(wvs))
abline(h = 139, col = "red")
```

### Interpretation of Variable Clusters:

Cluster 1: This cluster contains variables related to financial
satisfaction and democratic ideals, indicating an economic-political
ideology theme.

Cluster 2: Comprises variables focused on personal values, security, and
creativity, balancing traditional and innovative aspects.

Cluster 3: Primarily includes economic ideology questions such as
private vs. state ownership and competition, highlighting ideological
differences.

Cluster 4: Features variables emphasizing democratic values and
governance structures, particularly regarding the role of religion and
elections.
